# Visualizing Data {#sec-data-viz}

:::{.callout-note}
#### Packages Used
This chapter uses the following packages: `ggplot2`, `sjPlot`, `gtsummary`, `waffle`, `ggpubr`, 
`gridExtra`. 
:::

```{r, echo=FALSE}
source("_common.R")
library(waffle)
library(gridExtra)
```


Visualizing your data is hands down the most important thing you can learn to do. There are links to additional resources at the end of this document for additional learning. 

## Choosing your audience


The level of detail you put into your graphs/tables will depend on who
the graph/table is for. In general, you will be visualizing data for
three audience types: yourself and/or your data management team, an internal audience (other specialties within your organization, research coordinators), and external, professional audiences. The tabs below describe the three audience types and show an example graph for each.

**Who is your audience?**

::: panel-tabset
### Yourself

Graphs/tables for yourself and/or your working team can be "quick and dirty."
These graphs are for getting a quick look at the data and are meant for
people who are already familiar with the data.

```{r}
#| fig-height: 4
#| fig-weight: 3 
plot_frq(pen$island)

```

### Internal Audience

When presenting for an audience with some familiarity with the project and the data analysis process, your graphs don't have to be completely perfect, but they should be fairly polished and be understandable. A good rule of thumb for these graphs is if they
were to end up being published as-is without your knowledge, you
wouldn't be completely embarrassed to see them in print.

```{r}
#| fig-height: 4
#| fig-weight: 3 
plot_frq(pen$island, title = "Count of Penguins by Island") +
        xlab("Island") +
        theme_minimal()

```

### Professional

Professional quality graphs and tables have the highest amount of detail
and take the most amount of time to make. They should be able to be
interpreted by people not familiar with the project or data analysis, even
without reading the rest of the report.

```{r}
#| fig-height: 4
#| fig-weight: 3 
plot_frq(pen$island, title = "Count of Penguins by Island") +
          labs(subtitle = "Includes penguins of Adelie, Gentoo, and Chinstrap species") +
          xlab("Island") +
          ylab("Penguin Count") + 
          theme_minimal()

```
:::


## The Data

> switch to introduce the penguins dataset MAYBE use ### third level headers. 

We will use the `penguins` dataset that comes with the `palmerpenguins` package. This dataset contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antarctica. Review `?penguins` to learn about the variables we will be using. 


```{r}
pen <- palmerpenguins::penguins
```

> [RAD] write a paragraph here about how this section is organized. 

## One Categorical Variable

Both Nominal and Ordinal data types can be visualized using the same methods: tables, barcharts and pie charts. 


### Tables

Frequency and proporation (relative frequency) tables are the most common way to get summary statistics of a categorical variable.

::: {.panel-tabset}

## base

The `table()` function produces a frequency table, where each entry represents the number of records in the data set holding the corresponding labeled value. 

```{r}
table(pen$species)
```
There are 152 Adelie penguins, 68 Chinstrap penguins, and 124 Gentoo penguins in this dataset. 


## gtsummary

> Robin add text here. 

brief explanation of how it works. Defaults to creating a summary table of the entire data set, so you have to pre-select out only the variables you want to get a summary of. 

```{r}
pen |> select(species) |> tbl_summary()
pen |> select(species, island) |> tbl_summary()
```

This can be advantageous if you want to have a single table showing the frequency distribution of multiple variables. 
:::

### Frequency Barcharts / Barplots

::: {.callout-warning appearance=simple}

> adjust explanatory text here

:::  

A Barchart or barplot takes these frequencies, and draws bars along the X-axis where the height of the bars is determined by the frequencies seen in the table. 

::: {.panel-tabset}

## base

To create a barplot/barchart in base graphics requires the data to be in summarized in a table form first. Then the result of the table is plotted. The first argument is the table to be plotted, the `main` argument controls the title. 
```{r}
ps <- table(pen$species)
barplot(ps, main="Barchart using base graphics")
```

::: {.callout-warning appearance=simple}

> note yaxis direction and no labels at top of bar

:::  


## ggplot
The geometry needed to draw a barchart in ggplot is `geom_bar()`.
```{r}
ggplot(pen, aes(x=species)) + geom_bar()
```

**Adding Annotation**

The biggest addition to increase the readability of a barchart is to add the frequencies on top of the bars. 

```{r}
ggplot(pen, aes(x=species)) + theme_bw() + 
    geom_bar(aes(y = ..count..)) + ggtitle("Frequency of penguins by species") + 
    geom_text(aes(y=..count.. + 10, label=..count..), stat='count', size = 5)
```


## sjPlot
::: {.callout-warning appearance=simple}

> explain why this is the current best option

:::  

use sjplot example here. 


:::

### Proportion Barcharts

::: {.panel-tabset}

## ggplot

Often you don't want to compare counts but percents. To accomplish this, we have to aggregate the data to calculate the proportions first, then plot the aggregated data using `geom_col` to create the columns.

```{r}
spec.props <- data.frame(prop.table(table(pen$species)))
spec.props # what does this data look like? 

ggplot(spec.props, aes(x=Var1, y=Freq)) + geom_col() + 
  ylab("Proportion") + xlab("Species") + 
  ggtitle("Proportion of penguins by species")
```

## base

:::{.callout-warning appearance=simple}
* create a 'base' tab, but with only space for robin to write

:::

## sjplot

:::{.callout-warning appearance=simple}
* third tab is sjplot plot_frq (add this to the above section also)

:::

:::


### Other

> RAD add some intro text here. 

::: {.panel-tabset}
#### Cleveland Dot Plots

Another way to visualize categorical data that takes up less ink than bars is a Cleveland dot plot. Here again we are plotting summary data instead of the raw data. This uses the `geom_segment` that draws the lines from x=0 to the value of the proportion (named `Freq` because of the way `data.frame` works).

```{r}
ggplot(spec.props, aes(x=Freq, y=Var1)) +  
  geom_point(size = 3) + xlab("Proportion of penguins") + 
  theme_bw() + ylab("Species") +
  geom_segment(aes(x=0, xend=Freq, y=Var1, yend=Var1), color='grey50')
```

#### Pie Chart

Just like `barplot()`, `pie()` takes a table object as it's argument. 

**base**
```{r}
ps <- table(pen$species)
pie(ps)
```

Pie charts are my _least_ favorite plotting type. Human eyeballs can't distinguish between angles as well as we can with heights. A mandatory piece needed to make the wedges readable is to add the percentages of each wedge. 
```{r}
pie(ps, labels = paste0(names(ps), ' (', prop.table(ps)*100, "%)"))
```

**ggplot**

And here I thought pie charts couldn't get worse... I'm not a fan at all of the ggplot version. So i'm not even going to show it. [STHDA](http://www.sthda.com/english/wiki/ggplot2-pie-chart-quick-start-guide-r-software-and-data-visualization) has a great tutorial that does show you how to make one. 

However -- Never say never. [Storytelling with data](http://www.storytellingwithdata.com/blog/2019/8/8/forty-five-pie-charts-never-say-never) has a blog post with an example of a *good* use of pie charts. 


#### Waffle Chart 
This type of chart is not natively found in the `ggplot2` package, but it's own `waffle` package. These are great for infographics. 


```{r}
library(waffle)

waffle(ps/10, rows=5, size=0.5, 
       title="Species of penguins", 
       xlab="1 square == 10 penguins")
```

:::

## One Continuous Variable
Here we can look at the bill length and depth, flipper length, and body mass of the penguins. 

### Dotplot
> RAD change this to an introduction/motivating example instead of describing a plot type. 

```{r}
plot(pen$bill_length_mm)
```

The base function `plot()` creates a **dotplot** for a continuous variable. The value of the variable is plotted on the y axis, and the index, or row number, is plotted on the x axis. This gives you a nice, quick way to see the values of the data. 

Often you are not interested in the individual values of each data point, but the _distribution_ of the data. In other words, where is the majority of the data? Does it look symmetric around some central point? Around what values do the bulk of the data lie? 


### Histograms
Rather than showing the value of each observation, we prefer to think of the value as belonging to a \emph{bin}. **The height of the bars in a histogram display the frequency of values that fall into those of those bins.** For example if we cut the bill length distribution into 7 bins of equal width, the frequency table would look like this: 

```{r}
table(cut(pen$bill_length_mm, 7))
```

In a histogram, the binned counts are plotted as bars into a histogram. Note that the x-axis is continuous, so the bars touch. This is unlike the barchart that has a categorical x-axis, and vertical bars that are separated.

::: {.panel-tabset}

## base
You can make a histogram in base graphics super easy. 
```{r}
hist(pen$bill_length_mm)
```

And it doesn't take too much to clean it up. Here you can specify the number of bins by specifying how many `breaks` should be made in the data (the number of breaks controls the number of bins, and bin width) and use `col` for the fill color. 
```{r}
hist(pen$bill_length_mm, xlab="Bill length in mm", main="Histogram of penguin bill lengths", col="cyan", breaks=20)
```

## ggplot
```{r}
ggplot(pen, aes(x=bill_length_mm)) + geom_histogram(binwidth = 2.2)
```

The binwidth here is set by looking at the cut points above that were used to create 7 bins. Notice that darkgrey is the default fill color, but makes it hard to differentiate between the bars. So we'll make the outline black using `colour`, and `fill` the bars with white. 
```{r}
ggplot(pen, aes(x=bill_length_mm)) + geom_histogram(colour="black", fill="white") + 
  ggtitle("Distribution of penguin bill lengths")
```

Note I did **not** specify the `binwidth` argument here. The size of the bins can hide features from your graph, the default value for ggplot2 is range/30 and usually is a good choice. 

## ggpubr

> RAD add example here. 

:::

### Density plots
To get a better idea of the true shape of the distribution we can "smooth" out the bins and create what's called a `density` plot or curve. Notice that the shape of this distribution curve is much more... "wigglier" than the histogram may have implied. 

::: {.panel-tabset}

## base

```{r}
# plot(density(na.omit(pen$bill_length_mm))) 
pen$bill_length_mm |> na.omit() |> density() |> plot()
```

Awesome title huh? (NOT)

::: {.callout-warning appearance=simple}

> discuss why na.omit and nesting vs pipe example

:::  


## ggplot2

```{r}
ggplot(pen, aes(x=bill_length_mm)) + geom_density()
```

:::

### Histograms + density 
Often is is more helpful to have the density (or kernel density) plot _on top of_ a histogram plot. 

::: {.panel-tabset}

## base
Since the height of the bars in a histogram default to showing the frequency of records in the data set within that bin, we need to 1) scale the height so that it's a _relative frequency_, and then use the `lines()` function to add a `density()` line on top. 

```{r}
hist(pen$bill_length_mm, prob=TRUE)
lines(density(na.omit(pen$bill_length_mm)), col="blue")
```

## ggplot
The syntax starts the same, we'll add a new geom, `geom_density` and color the line blue. Then we add the histogram geom using `geom_histogram` but must specify that the y axis should be on the density, not frequency, scale. Note that this has to go inside the aesthetic statement `aes()`. I'm also going to get rid of the fill by using `NA` so it doesn't plot over the density line. 
```{r}
ggplot(pen, aes(x=bill_length_mm)) + geom_density(col="blue") + 
  geom_histogram(aes(y=..density..), colour="black", fill=NA)
```

:::

### Boxplots
Another very common way to visualize the distribution of a continuous variable is using a boxplot. Boxplots are useful for quickly identifying where the bulk of your data lie. R specifically draws a "modified" boxplot where values that are considered outliers are plotted as dots. 

::: {.panel-tabset}

## base
```{r}
boxplot(pen$bill_length_mm)
```

Notice that the only axis labeled is the y=axis. Like a dotplot the x axis, or "width", of the boxplot is meaningless here. We can make the axis more readable by flipping the plot on it's side. 
```{r}
boxplot(pen$bill_length_mm, horizontal = TRUE, main="Distribution of penguin bill lengths", xlab="Bill length in mm")
```

Horizontal is a bit easier to read in my opinion. 

## ggplot
What about ggplot? ggplot doesn't really like to do univariate boxplots. We can get around that by specifying that we want the box placed at a specific x value. 

```{r}
ggplot(pen, aes(x=1, y=bill_length_mm)) + geom_boxplot()
```

To flip it horizontal you may think to simply swap x and y? Good thinking. Of course it wouldn't be that easy. So let's just flip the whole darned plot on it's coordinate axis. 

```{r}
ggplot(pen, aes(x=1, y=bill_length_mm)) + geom_boxplot() + coord_flip()
```

:::

### Violin plots

```{r}
ggplot(pen, aes(x=1, y=bill_length_mm)) + geom_violin()
```

#### Boxplot + Violin plots
Overlaying a boxplot and a violin plot serves a similar purpose to Histograms + Density plots. 

```{r}
ggplot(pen, aes(x=1, y=bill_length_mm)) + geom_violin() + geom_boxplot()
```

Better appearance - different levels of transparency of the box and violin. 

```{r}
ggplot(pen, aes(x=1, y=bill_length_mm)) + xlab("") + theme_bw() + 
              geom_violin(fill="blue", alpha=.1) + 
              geom_boxplot(fill="blue", alpha=.5, width=.2) + 
              theme(axis.title.x=element_blank(),
              axis.text.x=element_blank(),
              axis.ticks.x=element_blank())
```



### Normal QQ plots
The last useful plot that we will do on a single continuous variable is to assess the _normality_ of the distribution. Basically how close the data follows a normal distribution. 

::: {.panel-tabset}

## base
```{r}
qqnorm(pen$body_mass_g)
qqline(pen$body_mass_g, col="red")
```

The line I make red because it is a reference line. The closer the points are to following this line, the more "normal" the shape of the distribution is. Price has some pretty strong deviation away from that line. Below I have plotted what a normal distribution looks like as an example of a "perfect" fit. 

```{r}
z <- rnorm(1000)
qqnorm(z)
qqline(z, col="blue")
```

## ggplot
qq (or qnorm) plots specifically plot the data against a theoretical distribution. That means in the `aes()` aesthetic argument we don't specify either x or y, but instead the `sample=` is the variable we want to plot. 
```{r}
ggplot(pen, aes(sample=body_mass_g)) + stat_qq()
```

:::


## Categorical v. Categorical

### Two-way tables

> intro to two way frequency and proportion tables

::: {.panel-tabset}

## frequency
Cross-tabs, cross-tabulations and two-way tables (all the same thing, different names) can be created by using the `table()` function. 

The frequency table is constructed using the `table()` function. 
```{r}
table(pen$island, pen$species)
```

There are 44 Adelie penguins and 124 Gentoo penguins on Biscoe Island, 56 Adelie and 68 Chinstrap penguins on Dream Island, and 52 Adelie penguins on Torgersen Island.

## proportion
Choose your percentages depending on your research question. What are you wanting to compare? 

Best practices: 

* Explanatory variable on the rows
* Response variable on the columns
* Calculate row %'s as the % of the response for each explanatory group. 

Here are demonstrations of how the interpretation of the percents change depending on what the denominator is. 

**Cell proportions**

Wrapping `prop.table()` around a table gives you the **cell** proportions. 
```{r}
prop.table(table(pen$island, pen$species))
```
12.8% of all penguins are Adelie penguins living on Biscoe Island, and 36% are Gentoo penguins living on Biscoe Island, so about 39% of all penguins in our data are living on Biscoe Island. 

**Row proportions**

To get the **row** proportions, you specify `margin=1`. The percentages now add up to 1 across the rows. 
```{r}
round(prop.table(table(pen$island, pen$species), margin=1),3)
```

26.2% of penguins on Biscoe Island are from the Adelie species. The remaining 73.8% on the island are Gentoo penguins.

**Column proportions**

To get the **column** proportions, you specify `margin=2`. The percentages now add up to 1 down the columns. 
```{r}
round(prop.table(table(pen$island, pen$species), margin=2),3)
```

0.289% of all Adelie penguins are on Biscoe Island. Chinstrap and Gentoo species are endemic to a single island, so 100% of their populations are on Dream and Torgersen Island, respectively.

:::

### Grouped bar charts
To compare proportions of one categorical variable within the same level of another, is to use grouped barcharts. 

::: {.panel-tabset}

## base
As before, the object to be plotted needs to be the result of a table. 
```{r}
is <- table(pen$island, pen$species)
barplot(is)
```

Stacked bars can be difficult to interpret, and very difficult to compare values between groups. A side by side barchart is preferable. 
The `beside=TRUE` is what controls the placement of the bars. 
```{r}
barplot(is, main="quick side by side barchart using base graphics", beside=TRUE)
```

## ggplot
Again plot the cut on the x axis, but then `fill` using the second categorical variable. This has the effect of visualizing the **row** percents from the table above, or the percent of species on each island. 

```{r}
ggplot(pen, aes(x=island, fill=species)) + geom_bar()
```

Again the default is a stacked barchart. So we just specify `position=dodge` to put the bars side by side. 
```{r}
ggplot(pen, aes(x=island, fill=species)) + geom_bar(position = "dodge")
```

And look, an automatic legend. What if I wanted to better compare island population proportions within species? This is the **column** percentages. Just switch which variable is the x axis and which one is used to fill the colors!
```{r}
ggplot(pen, aes(x=species, fill=island)) + geom_bar(position = "dodge")
```

For more than 2 colors I do not recommend choosing the colors yourself. I know little about color theory so I use the built-in color palettes. Here is a [great cheatsheet](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf) about using color palettes. 

And this easy change is why we love `ggplot2`. 

:::

### Grouped bar charts with percentages
Not as easy as one would hope, but the solution is to calculate the desired percentages first and then plot the summary data using either `geom_bar(stat='identity')` or `geom_col()`. 

```{r}
library(dplyr)
calc.props <- pen %>% group_by(island, species) %>%
              summarise(count=n()) %>%
              mutate(pct=round(count/sum(count),3))
calc.props
```

Since we're plotting summary data, the height of the bars is specified using `y=pct`. 

```{r}
ggplot(calc.props, aes(x=island, fill=species, y=pct)) +
                  geom_col(position="dodge") + theme_bw() 
```   

Now set some options to the y axis using `scale_y_continuous()` to make the graph more accurate and readable. The `labels=percent` comes from the `scales` package. 

```{r}
library(scales)
ggplot(calc.props, aes(x=island, fill=species, y=pct)) +
                  geom_col(position="dodge") + theme_bw() +
                  scale_y_continuous(limits=c(0,1), labels=percent)
```   



::: {.panel-tabset}

#### sjPlot

sjPlot does a very nice job of being able to cleanly show not only n's but percents. 
```{r}
library(sjPlot)
plot_xtab(pen$island, pen$species, margin="row", coord.flip = TRUE) 
```


#### Mosaic plots
But what if you want to know how two categorical variables are related and you don't want to look at two different barplots? Mosaic plots are a way to visualize the proportions in a table. So here's the two-way table we'll be plotting. 
```{r}
table(pen$island, pen$species)
```

The syntax for a mosaic plot uses _model notation_, which is basically y ~ x where the ~ is read as "twiddle" or "tilde". It's to the left of your **1** key.
```{r}
mosaicplot(island~species, data=pen)
```
> Increase fig length or use dif example variable

Helpful, ish. Here are two very useful options. In reverse obviousness, `color` applies shades of gray to one of the factor levels, and `shade` applies a color gradient scale to the cells in order of what is less than expected (red) to what is more than expected (blue) if these two factors were completely independent. 
```{r, fig.width=10}
par(mfrow=c(1,2)) # display the plots in 1 row and 2 columns
mosaicplot(island~species, data=pen, color=TRUE)
mosaicplot(island~species, data=pen, shade=TRUE)
```

For example, there are fewer 'Very Good' cut diamonds that are color 'G', and fewer 'Premium' cut diamonds that are color 'H'. As you can see, knowing what your data means when trying to interpret what the plots are telling you is essential. 

That's about all the ways you can plot categorical variables. 

If you are wondering why there was no 3D barcharts demonstrated see  
[here](http://faculty.atu.edu/mfinan/2043/section31.pdf),
[here](https://www.bbc.co.uk/bitesize/articles/z7cx7yc), and 
[here](https://en.wikipedia.org/wiki/Misleading_graph) for other ways you can really screw up your visualization.

:::

## Continuous v. Continuous 

### Scatterplot
The most common method of visualizing the relationship between two continuous variables is by using a scatterplot. 

::: {.panel-tabset}

## base
Back to the `plot()` command. Here we use model notation again, so it's $y~x$. 
```{r}
plot(flipper_length_mm~body_mass_g, data=pen)
```

Looks like for the most part as penguin body mass increases so does flipper length. That makes sense. 

## ggplot
With ggplot we specify both the x and y variables, and add a point. 
```{r}
ggplot(pen, aes(x=body_mass_g, y=flipper_length_mm)) + geom_point()
```

:::

### Adding lines to the scatterplots 

Two most common trend lines added to a scatterplots are the "best fit" straight line and the "lowess" smoother line. 

::: {.panel-tabset}

## base
The best fit line (in blue) gets added by using the `abline()` function wrapped around the linear model function `lm()`. Note it uses the same model notation syntax and the `data=` statement as the `plot()` function does. The lowess line is added using the `lines()` function, but the `lowess()` function itself doesn't allow for the `data=` statement so we have to use `$` sign notation. 

```{r}
plot(flipper_length_mm~body_mass_g, data=pen)
abline(lm(flipper_length_mm~body_mass_g, data=pen), col="blue")
lines(lowess(na.omit(pen$flipper_length_mm)~na.omit(pen$body_mass_g)),
      col="red")
```

## ggplot
With ggplot, we just add a `geom_smooth()` layer. 
```{r}
ggplot(pen, aes(x=body_mass_g, y=flipper_length_mm)) + geom_point() + geom_smooth() 
```

Here the point-wise confidence interval for this lowess line is shown in grey. If you want to turn the confidence interval off, use `se=FALSE`. Also notice that the smoothing geom uses a different function or window than the `lowess` function used in base graphics. 

Here it is again using the `ggplot` plotting function and adding another `geom_smooth()` layer for the `lm` (linear model) line in blue, and the lowess line (by not specifying a method) in red.

```{r}
ggplot(pen, aes(x=body_mass_g, y=flipper_length_mm)) + geom_point() + 
  geom_smooth(se=FALSE, method="lm", color="blue") + 
  geom_smooth(se=FALSE, color="red")
```

:::

### Line plots

Line plots connect each dot with a straight line. This is most often done when measuring trends of the response as the value of x increases (such as a time series)

We saw earlier that `body_mass_g` and `flipper_length_mm` seemed possibly linear. Let see how the average flipper length changes with body mass. 
```{r}
library(dplyr)
mm.per.g <- pen %>% group_by(body_mass_g) %>% summarise(mean = mean(flipper_length_mm))
```

::: {.panel-tabset}

## base
For base graphics, type='b' means both points and lines, 'l' gives you just lines and 'p' gives you only points. You can find more plotting character options under `?pch`. 

```{r}
plot(mean~body_mass_g, data=mm.per.g, type='l')
```

## ggplot
With ggplot we specify that we want a line geometry only. 
```{r}
ggplot(mm.per.g, aes(x=body_mass_g, y=mean)) + geom_line()
```

:::

How does this relationship change with penguin species? First lets 
get the average flipper length per combination of body mass and species 

```{r}
mmpgs <- pen %>% group_by(body_mass_g, species) %>% summarise(mean = mean(flipper_length_mm))
```

::: {.panel-tabset}

## ggplot
This is where ggplot starts to excel in it's ease of creating more
complex plots. All we have to do is specify that we want the lines 
colored by the cut variable. 

```{r}
ggplot(mmpgs, aes(x=body_mass_g, y=mean, col=species)) + geom_line()
```

And we get one line per cut. 

## base
This plot can be created in base graphics, but it takes an advanced 
knowledge of the graphics system to do so. So I do not show it here. 

:::

## Continuous v. Categorical
Create an appropriate plot for a continuous variable, and plot it for each
level of the categorical variable. 



### Dotplot/strip chart

Dotplots can be very useful when plotting dots against several categories. They can also be called stripcharts. 

::: {.panel-tabset}

## base
```{r}
stripchart(body_mass_g ~ species, data=pen)
```

## ggplot
We can reproduce the same thing by plotting one continuous variable against one categorical variable, and adding a layer of points. I'd argue that horizontal looks better due to the axis-labels. 
```{r}
library(gridExtra)
a <- ggplot(pen, aes(y=body_mass_g, x=species)) + geom_point()
b <- ggplot(pen, aes(y=species, x=body_mass_g)) + geom_point()
grid.arrange(a, b, ncol=2)
```
:::


### Grouped boxplots

::: {.panel-tabset}

## base

Base graphics plots grouped boxplots with also just the addition of a twiddle (tilde) `~`. 
Another example of where model notation works. 

```{r}
boxplot(bill_length_mm~species, data=pen)
```

## ggplot
A simple addition, just define your x and y accordingly. 
```{r}
ggplot(pen, aes(x=species, y=bill_depth_mm, fill=species)) + geom_boxplot()
```

## Adding violins

Violin plots can be overlaid here as well. 
```{r}
ggplot(pen, aes(x=species, y=bill_depth_mm, fill=species)) +
        geom_violin(alpha=.1) + 
        geom_boxplot(alpha=.5, width=.2)
```

:::

### Grouped histograms

::: {.panel-tabset}

## ggplot
By default ggplot wants to overlay all plots on the same grid. This doesn't look to good with histograms. Instead you can overlay density plots
```{r, fig.width=10}
a <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_histogram()
b <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density() 
grid.arrange(a,b, ncol=2)
```

The solid fills are still difficult to read, so we can either turn down the alpha (turn up the transparency) or only color the lines and not the fill. 
```{r, fig.width=10}
c <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density(alpha=.2)
d <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density() 
grid.arrange(c,d, ncol=2)
```

## base
There is no easy way to create grouped histograms in base graphics we will skip it. 

:::

### Joy plots / Ridgelines 

:::{.callout-warning appearance=simple}
check this section for updates - has this changed?

:::

Somewhat new (2017), joylines have not been added to the base distribution of `ggplot2` yet. For now it's available in the `ggjoy` package. Really good way to visualize density plots without the overlapping issue. 

```{r, fig.width=10}
library(ggjoy)
ggplot(pen, aes(x=bill_depth_mm, y=species)) + geom_joy()
```


## Faceting / paneling 

This is a good place to introduce a term called `faceting`. The definition is _a particular aspect or feature of something_, or _one side of something many-sided, especially of a cut gem_. Basically instead of plotting the grouped graphics on the same plotting area, we let each group have it's own plot, or facet.  

We add a `facet_wrap()` and specify that we want to panel on the species group. Note the twiddle in front of species 

```{r}
#| fig-height: 4
ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density() + facet_wrap(~species)
```

The grid placement can be semi-controlled by using the `ncol` argument in the `facet_wrap()` statement. 
```{r}
#| fig-height: 4

ggplot(pen, aes(x=bill_depth_mm, fill=species)) + 
  geom_density() + facet_wrap(~species, ncol=4)
```

It is important to compare distributions across groups on the same scale, and our eyes can compare items vertically better than horizontally. So let's force `ncol=1`. 

```{r}
#| fig-height: 4
#| 
ggplot(pen, aes(x=bill_depth_mm, fill=species)) + 
  geom_density() + facet_wrap(~species, ncol=1)
```


## Multiple plots per window

::: {.panel-tabset}

## base
I use `par(mfrow=c(r,c))` for base graphics, where `r` is the number of rows and `c` the number of columns. 

```{r}
par(mfrow=c(1,3))
plot(pen$bill_depth_mm)
plot(pen$species)
plot(pen$bill_length_mm ~ pen$bill_depth_mm)
```

Other resources including learning about `layouts`. [Multipanel plotting with base graphics](http://seananderson.ca/courses/11-multipanel/multipanel.pdf) 

## ggplot
Use the `grid.arrange` function in the `gridExtra` package. I've done it several times above. You assign the output of a ggplot object to an object (here it's `plot1` and `plot2`). Then you use `grid.arrange()` to arrange them either side by side or top and bottom. 
```{r}
a <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density(alpha=.2)
b <- ggplot(pen, aes(x=bill_depth_mm, fill=species)) + geom_density() 
grid.arrange(a,b, ncol=2)
```

## quarto

> example or reference to https://quarto.org/docs/get-started/computations/rstudio.html#multiple-figures 

:::

## Multivariate (3+ variables)
This is not much more complicated than taking an appropriate bivariate plot and adding a third variable through paneling, coloring, or changing a shape. 

This is trivial to do in ggplot, not trivial in base graphics. So I won't show those examples. 

### Three continuous
Continuous variables can also be mapped to the size of the point. Here I set the alpha on the points so we could see the overplotting (many points on a single spot). So the darker the spot the more data points on that spot. 

```{r}
ggplot(pen, aes(x=bill_length_mm, y=bill_depth_mm, size=body_mass_g)) + geom_point(alpha=.2)
```


### Scatterplot matrix
A scatterplot matrix allows you to look at the bivariate comparison of multiple pairs of variables simultaneously. First we need to trim down the data set to only include the variables we want to plot, then we use the `pairs()` function.

```{r}
c.vars <- pen[,c('flipper_length_mm', 'bill_length_mm', 'bill_depth_mm', 'body_mass_g')]
pairs(c.vars)
```
> Original comment- We can see price has a non-linear relationship with X, Y and Z and x & y have a near perfect linear relationship. 


### Two categorical and one continuous
This is very similar to side by side boxplots, one violin plot per `sex`, within each island 
> Original comment - This is difficult to really see due to the large number of categories each factor has. The categorical variables in the penguins dataset have less categories w less combinations 

```{r}
ggplot(pen, aes(x=island, y=bill_length_mm, fill=species)) + geom_violin()
```

Best bet here would be to panel on species and change the x axis to location.  
```{r}
ggplot(pen, aes(x=species, y=bill_length_mm, fill=species)) + geom_violin() + facet_wrap(~island)
```



### Two continuous and one categorical 
```{r}
a <- ggplot(pen, aes(x=bill_length_mm, y=bill_depth_mm, color=species)) + geom_point() + ggtitle("Colored by species")
d <- ggplot(pen, aes(x=bill_length_mm, y=bill_depth_mm, color=species)) + geom_point() + 
      geom_smooth(se=FALSE) +ggtitle("Lowess line per species")
grid.arrange(a, d, nrow=1)
```

Change the shape
```{r}
ggplot(pen, aes(x=bill_length_mm, y=bill_depth_mm, shape=species)) + geom_point() + ggtitle("Shape by species")
```

Or we just panel by the third variable

```{r}
ggplot(pen, aes(x=bill_length_mm, y=bill_depth_mm)) + geom_point() + facet_wrap(~species)
```

## Paneling on two variables
> move this up as a subsection of the faceting/paneling section 


Who says we're stuck with only faceting on one variable? A variant on `facet_wrap` is `facet_grid`. Here we can specify multiple variables to panel on. 

```{r, fig.width=10, fig.height=5}
ggplot(pen, aes(x=body_mass_g, fill=sex)) + geom_density() + facet_grid(sex~island)
```

How about plotting bill length against flipper length, for all combinations of species and island, with the points further separated by sex?
```{r}
ggplot(pen, aes(x=flipper_length_mm, y=bill_length_mm, color=sex)) + geom_point() + facet_grid(island~species)
```

And lastly let's look back at how we can play with scatterplots of using a third categorical variable (using `ggplot2` only). We can color the points by species, 

```{r}
ggplot(pen, aes(x=flipper_length_mm, y=bill_length_mm, color=species)) + geom_point()
```

We could add a smoothing lowess line for each species separately, 
```{r}
ggplot(pen, aes(x=flipper_length_mm, y=bill_length_mm, color=species)) + geom_point() + geom_smooth(se=FALSE)
```

We could change the color by sex, and shape by species. 
```{r}
ggplot(pen, aes(x=flipper_length_mm, y=bill_length_mm, color=sex, shape=species)) + geom_point() 
```

That's harder to read. So note that just because you **can** change an aesthetic, doesn't mean you should. And just because you can plot things on the same axis, doesn't mean you have to. 

Before you share your plot with any other eyes, always take a step back and try to explain what it is telling you. If you have to take more than a minute to get to the point then it may be too complex and simpler graphics are likely warranted. 


-----

## Troubleshooting 

### Problem: Missing data showing up as a category in ggplot? 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
NCbirths <- read.csv("https://www.norcalbiostat.com/data/NCbirths.csv", header=TRUE)
email <- read.table("https://www.norcalbiostat.com/data/email.txt", header=TRUE, sep="\t")
```

Get rid of that far right bar!
```{r}
ggplot(NCbirths, aes(x=marital)) + geom_bar()
```

::: {.callout-tip appearance=minimal collapse=true}

## Solution:
Use `dplyr` to select only the variables you are going to plot, then pipe in the `na.omit()` at the end. It will create a temporary data frame (e.g) `plot.data` that you then provide to `ggplot()`.

```{r}
plot.data <- NCbirths %>% select(marital) %>% na.omit()
ggplot(plot.data, aes(x=marital)) + geom_bar()
```

:::

### Problem: Got numerical binary 0/1 data but want to plot it as categorical? 

> Other related error messages: Continuous x aesthetic -- did you forget aes(group=...)?  

Consider a continuous variable for the number of characters in an email `num_char`, and a 0/1 binary variable `spam`. 

::: {.callout-tip appearance=minimal collapse=true}

## Solution:
Create a second variable `var_factor` for plotting and keep the binary `var` as 0/1 for analysis. 

```{r}
email$spam_cat <- factor(email$spam, labels=c("Ham", "Spam"))
ggplot(email, aes(y=num_char, x=spam_cat)) + geom_boxplot()
```

:::


### Problem: You want to change the legend title for a `fill` or `color` scale.  

::: {.callout-tip appearance=minimal collapse=true}

## Solution:
Add the `name=` argument to whatever layer you added that created the legend. Here I speciefied a `fill`, and it was a `discrete` variable. So I use the `scale_fill_discrete()` layer. 

```{r}
ggplot(email, aes(y=num_char, x=spam_cat, fill=spam_cat)) + geom_boxplot() + 
  scale_fill_discrete(name="Ya like Spam?")
```

Here, I `col`ored the points by a discrete variable, so the layer is `scale_color_discrete()`.
```{r}
ggplot(email, aes(x=num_char, y=line_breaks, col=spam_cat)) + geom_point() +
  scale_color_discrete(name="Ya like Spam?")
```

:::

### Problem: You want to add means to boxplots. 

Boxplots are great. Even better with violin overlays. Know what makes them even better than butter? Adding a point for the mean. `stat_summary` is the layer you want to add. Check out [this stack overflow post](https://stackoverflow.com/questions/23942959/ggplot2-show-separate-mean-values-in-box-plot-for-grouped-data) for more context. 

```{r}
ggplot(email, aes(x=spam_cat, y=num_char, fill=spam_cat)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", size=3, pch=17,color="blue")
```

::: {.callout-tip appearance=minimal collapse=true}

## Solution:

I suggest playing around with `size` and plotting character `pch` to get a feel for how these work. You can also look at `?pch` (and scroll down in the help file) to see the 25 default plotting characters.

:::


## But what about...

* Legend adjustment: remove it, move it to another side, rename it
* Custom specified colors and shapes

Go [here](http://www.cookbook-r.com/Graphs/) for these. 


* [Heat maps](https://www.r-bloggers.com/how-to-make-a-simple-heatmap-in-ggplot2/) 
* [Word clouds](https://rpubs.com/brandonkopp/creating-word-clouds-in-r), or [simpler](http://dangoldin.com/2016/06/06/word-clouds-in-r/)
* Interactive plots - Look into `plotly()` and `ggplotly()`, start with their guide directed towards R users [here](https://plotly.com/ggplot2/getting-started/) or browse the [Plotly R open source graphing library](https://plotly.com/r/)


## Additional Resources

For any Google Search -  be sure to limit searches to within the past year or so. R packages get updated very frequently, and many functions change or become obsolete. 

* Comprehensive guides
  - [R Graphics 2e](https://www.stat.auckland.ac.nz/~paul/RG2e/): The best book about using base graphics
  - [R Graphics Cookbook](https://r-graphics.org): The best book for using ggplot2 
  - [ggplot2: Elegant Graphics for Data Analysis 3e](http://ggplot2.org/book/) 

* Subject reference and tutorial sites
  - [Math 130 Lesson 07 - Creating graphics](https://norcalbiostat.github.io/MATH130/notes/07_plots.html#The_syntax_of_ggplot) - more on ggplot2 syntax
  - [STHDA](http://www.sthda.com/english/): Statistical tools for high-throughput data analysis.
    - [ggplot2 essentials](http://www.sthda.com/english/wiki/ggplot2-essentials)
    - [Other graphics options](http://www.sthda.com/english/articles/32-r-graphics-essentials/)
    - [ggplot2 qq plot (quantile - quantile graph) : Quick start guide](http://www.sthda.com/english/wiki/ggplot2-qq-plot-quantile-quantile-graph-quick-start-guide-r-software-and-data-visualization)
  -[Quick-R / Datacamp](https://www.statmethods.net): 
    - [Basic Graphs](http://www.statmethods.net/graphs/index.html)
    - [ggplot2](http://www.statmethods.net/advgraphs/ggplot2.html)
    - [Scatterplots](http://www.statmethods.net/graphs/scatterplot.html)  
  - [R-bloggers](https://www.r-bloggers.com)
    - [Waffle charts](https://www.r-bloggers.com/making-waffle-charts-in-r-with-the-new-waffle-package/)
    - [Scatterplot matrices](https://www.r-bloggers.com/scatterplot-matrices/)

* Literate programming tools
  - Quarto
    - [DMultiple figure alignment](https://quarto.org/docs/get-started/computations/rstudio.html#multiple-figures) 

* Help lists 
    - [ggplot2 Google Groups](http://groups.google.com/group/ggplot2) 
    - [Stackoverflow questions tagged ggplot2](http://stackoverflow.com/tags/ggplot2)
    - [Chico R users group](https://www.meetup.com/chico-r-users-group/))

::: {.callout-tip}
### Learn more
For more practice with `ggplot` syntax, reference 

:::